### 2020-11-20 pldi21/case-studies.tex
ported to PLDI format and tightened up abstract
reworking intro
tighten up sec 2

\section{Case Studies}\label{sec:case-studies}
Let us begin by considering livelits from the client programmer's perspective by way of
two domain-specific case studies:
a course grade assignment case study in Sec.~\ref{sec:live-grading}
and an image transformation case study in Sec.~\ref{sec:image-transformation}.
% We also briefly mention several other examples in Sec.~\ref{sec:additional-examples}.\todo{do we?}{}
These case studies have been implemented
in Hazel, a browser-based live programming environment for a pure typed functional language in the
ML family. We assume basic familiarity with ML.

\subsection{Case Study: Grading with Livelits}\label{sec:live-grading}
\begin{figure}
    \includegraphics[width=30pc]{grade-cutoff-livelit.png}
\caption{Grading with Livelits}
\label{fig:grading}
\end{figure}

Consider this familiar scenario: an instructor needs
(1) to record numeric grades for various assignments and exams, and
(2) to visualize and perform various computations with these numeric grades
in order ultimately to assign final letter grades (\textbf{A} through \textbf{F}) to the students in the course.

Perhaps the most common end-user application in this domain is the spreadsheet.
A spreadsheet allows the instructor to record student grades using a natural tabular interface,
visualize this data in one of a finite number of plot styles, and perform basic computations on this data,
with results updated live, i.e. automatically after each cell has been edited.
However, these affordances are limited. It is difficult to package up common operations and workflows
into reusable libraries, interact with the data using domain-specific visualizations,
and perform operations that have not been anticipated by the designers of the spreadsheet system
(e.g. operations that prepare the data in an idiosyncratic format suitable for automated submission
to the university's grading system).
General-purpose programming languages equipped with suitable libraries for working with numeric data
are flexible enough to handle these scenarios, but users of these languages
sacrifice the ability to receive live feedback and directly manipulate data and visualizations thereof.

Livelits are able to address this tension.
Fig.~\ref{fig:grading} shows a Hazel program where the user has:
\begin{enumerate}
\item recorded numeric \li{grades} for each student using a livelit, \li{\$dataframe}, that implements a tabular user interface with support for live evaluation of the Hazel expression in each cell,
\item programmatically computed the overall \li{averages} for each student by applying a function, \li{compute_averages}, defined in a library  (not shown) shared between multiple courses,
\item determined reasonable \li{cutoffs} between letter grades by directly manipulating a livelit that offers a domain-specific drag-and-drop interface superimposed on a live visualization of the grade distribution
(for the sake of space, we plot dots rather than a full histogram),
\item programmatically assigned letter grades to each student based on the \li{cutoffs} determined in step (3) by calling another shared library function, \li{assign_grades},
\item formatted the final grade assignments in a manner suitable for submission to the university system by calling another shared library function, \li{format_for_university}.
\end{enumerate}

\subsubsection{Livelit Expansion}\label{sec:livelit-expansion}
Livelit expressions are given dynamic meaning by expansion to core expressions, i.e. expressions that do not contain livelits.
For example, the expansion of Fig.~\ref{fig:grading} is:

\begin{lstlisting}
let grades = Dataframe (
  ["A1", "A2", "A3", "A4", "Midterm", "Final"],
  [
    ("Alice", [24. +. 36. +. 33., 92., 83.5, 87.5, 95., 88.]),
    ("Bob", [61., 64., 98., 89.5, 70., 85.]),
    ("Carol", [75., 81., 73., 71., 82., 79.]),
    (* ... *)
  ]) in
let averages = compute_averages grades weights in
let cutoffs = (.A 90.0, .B 80.0, .C 70.0, .D 60.0) in
format_for_university (assign_grades averages cutoffs)
\end{lstlisting}

>>>>

\section{Livelits by Example}\label{sec:case-studies}

In this section, we will detail the livelits mechanism by way of  
two domain-specific case studies:
a course grade assignment case study in Sec.~\ref{sec:live-grading}
and an image transformation case study in Sec.~\ref{sec:image-transformation}.
% We also briefly mention several other examples in Sec.~\ref{sec:additional-examples}.\todo{do we?}{}
These case studies have been implemented
in Hazel, a browser-based live programming environment for a dialect of Elm. 
Elm is an industrial pure typed functional language in the
ML family used for client-side web development. 
We assume basic familiarity with ML.

\subsection{Case Study: Grading with Livelits}\label{sec:live-grading}
Consider this familiar scenario: an instructor needs
(1) to record numeric grades for various assignments and exams, and
(2) to visualize and perform various computations with these numeric grades
in order ultimately to assign final letter grades.
(In fact, this case study is not contrived: one author is using Hazel to compute grades this semester.)

The most common end-user application for this task is the spreadsheet, because 
it allows the instructor to record grades using a natural tabular interface,
visualize this data in one of a finite number of plot styles, 
and perform basic computations,
with results updated live.
However, these affordances are limited. 
It is difficult to package up common operations 
into reusable libraries, interact with the data using domain-specific visualizations,
and perform complex, unanticipated operations 
(e.g. preparing the data in an idiosyncratic format demanded by the university's grading system).

General-purpose programming languages 
can handle these scenarios, but users 
lose the ability to receive live feedback and  
directly manipulate data and visualizations in the editor.

Livelits are able to address this tension.
Fig.~\ref{fig:grading}(c)\todo{sublabels} shows a Hazel program where 
the instructor alternates between programmatic and direct manipulation in several situations.

First, the instructor defines a value \li{grades} 
that records the grades for each student using a livelit, \li{\$dataframe}, 
that implements a tabular user interface. The formula bar 
allows the selected cell to be filled with an arbitrary Hazel expression, 
with all of Hazel's editor services available. 
For the sake of demonstration, we show a cell that has been filled using another livelit, 
\li{\$slider}, in combination with symbolic manipulation.\todo{do this?}{}
The table itself displays not the expression itself but rather its value, just as in a spreadsheet.

Next, the instructor programmatically computes the overall \li{averages} 
for each student by applying \li{compute_averages}, a helper function 
defined in the usual way in a library  (not shown) shared between multiple courses.

Next, the instructor wants to ``eyeball'' reasonable \li{cutoffs} between letter grades 
by directly manipulating a highly domain-specific livelit, \li{\$cutoffs}, that provides draggable ``paddles'' 
superimposed on a live visualization of the distribution of the \li{averages} value provided as a parameter 
(for the sake of space, we plot dots rather than a histogram).
% The value of \li{cutoffs} is a labeled 4-tuple containing each cut-off.

Finally, 
the instructor programmatically assigns letter grades to each student 
based on the selected \li{cutoffs} 
by calling \li{assign_grades} 
and formats these for submission with \li{format_for_university},
again shared functions.

\subsubsection{Livelit Expansion}\label{sec:livelit-expansion}
Livelit expressions are given dynamic meaning by expansion to core expressions, which do not contain livelits.
For example, the expansion of Fig.~\ref{fig:grading} is:

\begin{lstlisting}[xleftmargin=0.2cm]
let grades = Dataframe (
  ["A1", "A2", "A3", "Midterm", "Final"],
  [("Alice", [24. +. 36. +. 33., 
               92., 83.5, 95., 88.]),
   ("Bob", [61., 64., 98., 70., 85.]),
   ("Ciri", [75., 81., 73., 82., 79.]),
   (* ... *) ]) in
let averages = compute_averages grades weights in
let cutoffs = (.A 90., .B 80., .C 70., .D 60.) in
format_for_university 
  (assign_grades averages cutoffs)
\end{lstlisting}

### 2020-11-20 pldi21/intro.tex
tighten up sec 2

\llproperty{Liveness} Graphite's palettes do not have
  access to the run-time environment. By contrast, livelits can evaluate splices
  and therefore provide feedback related to the run-time behavior of the expression being generated.
  For example, in Fig.~\ref{fig:color}(b), the color that is displayed is determined by evaluating the RGBA
  component splices to numeric values.
  Evaluation occurs in a run-time environment (i.e. closure) determined by
  leaving the hole being filled by the livelit temporarily unfilled and then evaluating
  using a two-phased variation on the semantics for incomplete programs developed for the \Hazel programming environment \cite{HazelnutLive}.
  We will see more sophisticated examples later on, including a livelit
  that appears inside a function that is called multiple times, leading to multiple closures that the client can
   select between.

\paragraph{Outline.} We begin in Sec.~\ref{sec:case-studies} from the livelit client perspective 
with case studies involving several non-trivial livelits chosen to demonstrate the novel contributions of this paper.
In Sec.~\ref{sec:livelit-definitions}, we consider the livelit provider's perspective by detailing the livelit
definition mechanism by example.
In Sec.~\ref{sec:livelit-calculus}, we provide a formal account of the livelit mechanism, including structural representation, expansion, and live closure collection, as a typed lambda calculus. 
We have mechanically specified the central mechanism, livelit expansion, and proven the associated metatheorems in Agda.
The livelit calculus 
serves to capture the essential nature of the mechanism independent of the particularities of syntax, a GUI framework, and other irrelevant design and implementation details,
because we believe livelits can be integrated into a wide variety of programming systems. 
In Sec.~\ref{sec:implementation}, we provide a more detailed account of our two implementations of livelits.
Our primary implementation is integrated into Hazel, a live programming environment designed from the ground up 
around live hole-driven development. 
We have also prototyped livelits within a more conventional text editor 
and discuss several considerations unique to this setting.
In Sec.~\ref{sec:related-work}, we compare livelits to related work using the design criteria outlined above.
Finally, we conclude in Sec.~\ref{sec:discussion} with a discussion of some present limitations and future work.

>>>>

\llproperty{Liveness} Uniquely, livelits can evaluate splices
  and so provide feedback related to the run-time behavior of the expression being generated.
  For example, in Fig.~\ref{fig:color}(b), 
  the color that is displayed is determined by evaluating the RGBA
  component splices to numeric values.
  Evaluation occurs in a run-time environment (i.e. closure) determined by
  leaving the hole being filled by the livelit temporarily unfilled and then evaluating
  using a two-phased variation on the semantics for 
  live programming with typed holes developed by \citet{HazelnutLive}.
  We support live evaluation even for livelits 
  that appears inside a function that is called multiple times: this leads 
  to multiple closures that the client 
   selects between.

\paragraph{Outline.} We begin in Sec.~\ref{sec:case-studies} by introducing
livelits from the perspective of client programmers. 
Our examples are organized into case studies 
and are chosen to specifically demonstrate the novel contributions of this paper.
In Sec.~\ref{sec:livelit-definitions}, 
we consider the livelit provider's perspective by detailing the livelit
definition mechanism with an example.
In Sec.~\ref{sec:livelit-calculus}, we define the \emph{typed livelit calculus}. 
We have mechanically specified the central mechanism, livelit expansion, and proven the associated metatheorems in Agda.
This calculus 
serves to capture the essential nature of livelits 
independent of the particularities of syntax, GUI frameworks, 
and other orthogonal design details,
because we believe livelits can be integrated into a wide variety of programming systems. 
In Sec.~\ref{sec:implementation}, we provide a more detailed account of our two implementations of livelits.
Our primary implementation, used in the screenshots in the paper,
is within Hazel, a live programming environment designed 
around hole-driven development. 
We have also prototyped livelits within a standard text editor. 
We also discuss factors to consider when adding livelits to languages with side effects.
In Sec.~\ref{sec:related-work}, we compare livelits to related work using the design properties outlined above 
as a rubric.
Finally, we conclude in Sec.~\ref{sec:discussion} after a discussion of some present limitations and future work.
